---
title: "mRNA plots"
---
THIS IS THE LATEST EDITION WITH THE FINAL PLOTS. 
Supplemental plots may be in one of the other qPCR/*.qmd files

All data cleaning, tidying and wrangling was done before the start of this quarto manuscript project. Code is in: C:/Users/alosvendsen/OneDrive - Syddansk Universitet/PhD/R_code/manuscript_1 And copied to this project: /notebooks/qPCR/old_qPCR_import

Donor 1: timepoints = 72 and 120 hours Donor 2: timepoints = 24, 72, and 120 hours Donor 2: timepoints = 24, 72, and 120 hours

## Setup

```{r}
#| warning: false
#| message: false

library(here)
library(data.table)
library(ggplot2)
library(patchwork)
library(RColorBrewer)
library(viridis)
source(here("notebooks/qPCR/custom_functions.R"))
```

## Finalize the data and save the data table
Just done once, and commented out hereafter. 

### Load data, remove 0.001 ng/ml cytokine concentrations, calculate means for each experiment.
Already done. Commented out. Final plot data i ./data_final_for_plotting/
```{r}
#### Load data, filter data and calculate geometric means
# data <- readRDS(here("notebooks/qPCR/data_processed/final_dataframe.rds"))
#
# # Remove the lowest concentrations of cytokines, since not used in all
# # Experiments. Note 24 hour time points was only used in one experiment
# plot_data <- data[!treatment %chin% c("0.001 ng/ml IL-1B", "0.001 ng/ml IL-6")]
#
#
# # Calculate the mean expression grouped by experiment, condition, treatment, and time.
# # USE GEOMETRIC MEAN when averaging normalized expression (RQ) values, since they are on a multiplicative scale.
# plot_data[,
#           bio_group_mean_expression := geometric_mean(normalized_expression),
#           by = .(experiment, condition, treatment, time, target)
#           ]
#
# # Save the data
# saveRDS(plot_data, here("notebooks/qPCR/data_final_for_plotting/plot_data.rds"))
```

```{r}
# rmv <- ls()[!grepl(x =ls(), pattern = "geometric")]
#
# rm(list = rmv)
```

## Main plot

Remember too use geometric mean when on a multiplicative scale (such as when handling RQ/fold data!)
```{r}
# Load the data
plot_data <- readRDS(
    here("notebooks/qPCR/data_final_for_plotting/plot_data.rds")
)

# Filter for CYP enzymes only, include all treatments and CYPs
cyp_data <- plot_data[target %like% "CYP"]

# Split data based on treatment condition
cyp_data_IL6 <- cyp_data[condition == "IL-6"]
cyp_data_IL1B <- cyp_data[condition == "IL-1B"]

# calc_means <- function(cyp_data) {
#     cyp_data[,
#     .(mean_normalized_expression = geometric_mean(normalized_expression)),
#     by = .(time, treatment, target)
#     ]
# }
# calc_means(cyp_data_IL6)

# Create plot function. Takes data, and condition arguments.
create_plot_with_time_grouping <- function(cyp_data, condition) {
    # Calculate mean normalized expression for each combination of:
    # CYP enzyme, treatment and time
    cyp_means <-
        cyp_data[,
            .(
                mean_normalized_expression =
                    geometric_mean(normalized_expression)
            ),
            by = .(time, treatment, target)
        ]

    # Calculate the overall mean across time points
    cyp_overall_means <-
        cyp_means[,
            .(
                overall_mean_normalized_expression =
                    geometric_mean(mean_normalized_expression)
            ),
            by = .(treatment, target)
        ]

    # Create and return the plot
    ggplot() +
        # Plot mean of each timepoint
        geom_point(
            data = cyp_means,
            aes(
                x = treatment,
                y = mean_normalized_expression,
                fill = factor(time),
                shape = factor(time)
            ),
            alpha = 0.7,
            size = 7,
            stroke = 1,
            color = "black"
        ) +
        # Plot overall mean
        geom_point(
            data = cyp_overall_means,
            aes(
                x = treatment,
                y = overall_mean_normalized_expression
            ),
            color = "black",
            size = 40,
            shape = 45, # Black horizontal rectangle for overall mean
            alpha = 0.85
        ) +
        geom_hline(yintercept = 1, linetype = "dashed") +
        scale_y_continuous(limits = c(0, NA)) +
        # Modify shape to include the overall mean
        scale_shape_manual(
            values = c(21, 22, 24, 45), # Circle, Triangle, Square, Horizontal Rectangle
            labels = c("24", "72", "120", "Overall Mean") # Add "Overall Mean" label
        ) +
        scale_fill_viridis(discrete = TRUE, option = "viridis", begin = 0, end = .9) + # Using viridis color scale
        facet_wrap(~target, scales = "free_y", ncol = 2) +
        labs(
            x = "Treatment", y = "Mean Normalized Expression",
            title = paste("Normalized Expression after", condition),
            fill = "Time (hours)",
            shape = "Time (hours)" # Shared label for timepoints and overall mean
        ) +
        theme_bw() +
        theme(
            text = element_text(size = 14), # Base font size
            axis.text = element_text(size = 12),
            axis.title = element_text(size = 16, face = "bold"),
            axis.text.x = element_text(angle = 45, hjust = 1),
            plot.title = element_text(size = 20, face = "bold"),
            legend.title = element_text(size = 14, face = "bold"),
            legend.text = element_text(size = 12),
            legend.key.size = unit(2, "lines"), # Increase size of keys in the legend
            legend.spacing = unit(0.8, "cm"), # Increase spacing between legend items
            strip.background = element_rect(fill = "lightgrey"),
            strip.text = element_text(size = 16, face = "bold")
        )
}


# Create the plots
plot_IL6 <- create_plot_with_time_grouping(cyp_data_IL6, "IL-6")
plot_IL1B <- create_plot_with_time_grouping(cyp_data_IL1B, "IL-1β")
plot_IL1B
plot_IL6

# Save the plots
ggsave(
    filename = "images/mRNA/plots/main_normalized_Expression_IL6.png",
    plot = plot_IL6,
    device = "png",
    width = 12,
    height = 10,
    dpi = "retina"
)

ggsave(
    filename = "images/mRNA/plots/main_normalized_Expression_IL1B.png",
    plot = plot_IL1B,
    device = "png",
    width = 12,
    height = 10,
    dpi = "retina"
)
```

